---
title: "Flights and Departure Delays"
author: "Jonah Maroszek"
date: "2/19/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(nycflights13)
```

# Introduction 
I chose to examine how precipitation, age of the plane, and destination would affect the average departure delay. In summary:  

* Precipitation and departure delay does not follow a linear relationship as I would have expected, but in general more precipitation results in a greater departure delay.  
* The age of the plane and departure delay are positively related up until 10 years, and then negatively related until the planes are taken out of commission. This also goes against what I would have expected.  
* NEVER fly to Oklahoma (unless you want to experience long delays)

### Impact of Precipitation on Average Delay
I expected there to be clear linear relationship between the amount of precipitation and the average delay. It appears that this relationship does not hold in general, as there are oscillations. However, as a rough estimate, lower precipitation typically results in a shorter departure delay
```{r Weather, echo=FALSE, message=FALSE, warning=FALSE}
flight_weather <-
  flights %>%
  inner_join(weather, by = c(
    "origin" = "origin",
    "year" = "year",
    "month" = "month",
    "day" = "day",
    "hour" = "hour"
  ))

flight_weather %>%
  group_by(precip) %>%
  summarize(delay = mean(dep_delay, na.rm = TRUE)) %>%
  ggplot(aes(x = precip, y = delay)) +
  geom_line() + 
  geom_point() +
  xlab("Precipitation") +
  ylab("Mean Departure Delay (Minutes)")

```


### Impact of Age of Plane on Average Delay
There is an inverted U shape when you plot average departure delay versus the age of the plane, with the peak occuring at around 10 years. This also defies my expectation. It looks like if you want to experience the lowest delay, you should fly on either a very young or very old plane. The trend could reverse at around 10 years because the airlines keep a closer eye on the possible mechanical issues that could cause delays, and either remove them from service or make sure there is enough time to fix them.  
```{r Age of Plane, echo=FALSE, message=FALSE, warning=FALSE}
planeStuff <- inner_join(flights,
  select(planes, tailnum, plane_year = year), by = "tailnum") %>%
  mutate(age = year - plane_year) %>% #calculate how old plane is 
  filter(!is.na(age)) %>%
  mutate(age = if_else(age > 25, 25L, age)) %>% #focus on planes younger than 25 years
  group_by(age) %>%
  summarize(
    dep_delay_mean = mean(dep_delay, na.rm = TRUE),
  )

ggplot(planeStuff, aes(x = age, y = dep_delay_mean)) +
  geom_point() +
  scale_x_continuous("Age of Plane (years)", breaks = seq(0, 30, by = 10)) +
  scale_y_continuous("Mean Departure Delay (minutes)") +
  geom_smooth(se = F)


```


### Impact of Destination on Average Delay 
If you are flying to a city designated by a dark purple circle, you should expect a higher average delay than the lighter color cities.
```{r Destination Delay, echo=FALSE, message=FALSE, warning=FALSE}
mean_dest_delay <- flights %>% 
  group_by(dest) %>% 
  summarize(mean_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  inner_join(airports, by = c(dest = "faa"))

mean_dest_delay %>% 
  ggplot(aes(lon,lat, color = mean_delay)) +
  borders("state") + 
  geom_point() +
  coord_quickmap() +
  xlab("longitude") +
  ylab("latitude") +
  ggtitle("Average Delay by Destination") +
  scale_color_gradient2() 
  
```

